---
title: "Spilhaus, as it flows downstream from PROJ: the R view"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
execute:
  error: true
  warning: false
  message: false
---

The companion to [the Python notebook](../index.html). Same sources, same
sections, same landmarks, different runtime. The point is the same too:
`+proj=spilhaus` arrived in **PROJ 9.6.0** (March 2025), and whether *this*
R session can use it depends on which PROJ each package was linked against.

R's version of the problem is not the Python one. On Linux, sf, terra,
gdalraster and vapour all link one system `libproj`, so they normally agree;
the question is what the *system* has, and on a stock Ubuntu that is often
too old. On macOS and Windows, CRAN's binaries bundle their own GDAL and
PROJ, chosen by CRAN's build recipes rather than by package authors. So the
row that changes between machines is the operating system, not the wheel.

The stack here is deliberately low-level: **gdalraster** for I/O and the warp,
**wk** for vertex-level geometry, **ximage** to draw a matrix. sf and terra
appear only in the provenance table and in an appendix, as the same operation
in one line, so you can see which PROJ each one silently used.

```{r setup}
SPILHAUS <- "+proj=spilhaus"
SPILHAUS_EXPLICIT <- paste(
  "+proj=spilhaus +lon_0=66.94970198 +lat_0=-49.56371678",
  "+azi=40.17823482 +rot=45 +k_0=1 +x_0=0 +y_0=0 +R=6378137"
)
SPILHAUS_ALT <- "+proj=spilhaus +lon_0=-30 +lat_0=-20 +azi=20 +rot=45"

GEBCO_URL <- "/vsicurl/https://data.source.coop/alexgleith/gebco-2024/GEBCO_2024.tif"
RASTER_URL <- Sys.getenv("SPILHAUS_RASTER", GEBCO_URL)

NE_OCEAN_URL <- paste0(
  "/vsicurl/https://raw.githubusercontent.com/nvkelso/natural-earth-vector/",
  "master/110m_physical/ne_110m_ocean.shp"
)
OCEAN_URL <- Sys.getenv("SPILHAUS_OCEAN", NE_OCEAN_URL)

NPIX <- as.integer(Sys.getenv("SPILHAUS_NPIX", "1600"))

Sys.setenv(GDAL_DISABLE_READDIR_ON_OPEN = "EMPTY_DIR",
           GDAL_HTTP_MULTIRANGE = "YES",
           GDAL_HTTP_MERGE_CONSECUTIVE_RANGES = "YES")
```

## 1. Who has which PROJ?

One row per package. `spilhaus` is the result of asking that package to make
a CRS from `+proj=spilhaus`.

```{r provenance}
probe <- function(expr) {
  out <- tryCatch({ force(expr); "ok" }, error = function(e) paste("FAIL:", conditionMessage(e)))
  substr(out, 1, 80)
}
has <- function(pkg) requireNamespace(pkg, quietly = TRUE)

rows <- list()

if (has("gdalraster")) {
  pv <- gdalraster::proj_version()
  rows$gdalraster <- data.frame(
    package = "gdalraster", version = as.character(packageVersion("gdalraster")),
    gdal = gdalraster::gdal_version()[4], proj = pv$name,
    spilhaus = probe(gdalraster::srs_to_wkt(SPILHAUS)))
} else rows$gdalraster <- data.frame(package = "gdalraster", version = NA, gdal = NA, proj = NA, spilhaus = "not installed")

if (has("PROJ")) {
  rows$PROJ <- data.frame(
    package = "PROJ", version = as.character(packageVersion("PROJ")),
    gdal = "-", proj = PROJ::proj_version(),
    spilhaus = probe(PROJ::proj_trans(cbind(147.33, -42.88), SPILHAUS, source = "OGC:CRS84")))
} else rows$PROJ <- data.frame(package = "PROJ", version = NA, gdal = "-", proj = NA, spilhaus = "not installed")

if (has("sf")) {
  v <- sf::sf_extSoftVersion()
  rows$sf <- data.frame(
    package = "sf", version = as.character(packageVersion("sf")),
    gdal = unname(v["GDAL"]), proj = unname(v["PROJ"]),
    spilhaus = probe({ x <- sf::st_crs(SPILHAUS); if (is.na(x)) stop("st_crs returned NA") }))
} else rows$sf <- data.frame(package = "sf", version = NA, gdal = NA, proj = NA, spilhaus = "not installed")

if (has("terra")) {
  rows$terra <- data.frame(
    package = "terra", version = as.character(packageVersion("terra")),
    gdal = terra::gdal(lib = "gdal"), proj = terra::gdal(lib = "proj"),
    spilhaus = probe({ x <- terra::crs(SPILHAUS); if (!nzchar(x)) stop("empty crs") }))
} else rows$terra <- data.frame(package = "terra", version = NA, gdal = NA, proj = NA, spilhaus = "not installed")

if (has("vapour")) {
  rows$vapour <- data.frame(
    package = "vapour", version = as.character(packageVersion("vapour")),
    gdal = vapour::vapour_gdal_version(), proj = vapour::vapour_proj_version(),
    spilhaus = probe(vapour::vapour_srs_wkt(SPILHAUS)))
}

if (has("wk")) rows$wk <- data.frame(package = "wk", version = as.character(packageVersion("wk")),
                                     gdal = "-", proj = "- (no PROJ dependency)", spilhaus = "n/a")
if (has("ximage")) rows$ximage <- data.frame(package = "ximage", version = as.character(packageVersion("ximage")),
                                             gdal = "-", proj = "-", spilhaus = "n/a")

provenance <- do.call(rbind, rows); rownames(provenance) <- NULL
knitr::kable(provenance, caption = sprintf("R %s on %s", getRversion(), R.version$platform))
```

```{r cli}
cli_version <- function(cmd, args = "--version") {
  exe <- Sys.which(cmd)
  if (!nzchar(exe)) return("not found")
  out <- tryCatch(system2(exe, args, stdout = TRUE, stderr = TRUE), error = function(e) conditionMessage(e))
  out[1]
}
cli <- c(gdalinfo = cli_version("gdalinfo"),
         gdal = cli_version("gdal"),
         projinfo = cli_version("projinfo"))
knitr::kable(data.frame(tool = names(cli), version = unname(cli)), caption = "Command-line tools on PATH")
```

```{r engine}
capable <- provenance$package[provenance$spilhaus == "ok" & provenance$package %in% c("gdalraster", "PROJ", "sf", "terra", "vapour")]
if (length(capable) == 0) stop("No installed package can create +proj=spilhaus on this machine; nothing below can run.")
ENGINE <- capable[1]
cat("Coordinate transforms below use:", ENGINE, "\n")

## forward/inverse constructors for any proj string, from the first capable package
make_forward <- function(proj_string) {
  switch(ENGINE,
    gdalraster = function(lonlat) gdalraster::transform_xy(lonlat, "OGC:CRS84", proj_string),
    PROJ       = function(lonlat) PROJ::proj_trans(lonlat, proj_string, source = "OGC:CRS84"),
    sf         = function(lonlat) sf::sf_project("OGC:CRS84", proj_string, lonlat, keep = TRUE),
    terra      = function(lonlat) terra::crds(terra::project(terra::vect(lonlat, crs = "OGC:CRS84"), proj_string)),
    vapour     = function(lonlat) vapour::vapour_project(lonlat, proj_string, "OGC:CRS84"))
}
make_inverse <- function(proj_string) {
  switch(ENGINE,
    gdalraster = function(xy) gdalraster::transform_xy(xy, proj_string, "OGC:CRS84"),
    PROJ       = function(xy) PROJ::proj_trans(xy, "OGC:CRS84", source = proj_string),
    sf         = function(xy) sf::sf_project(proj_string, "OGC:CRS84", xy, keep = TRUE),
    terra      = function(xy) terra::crds(terra::project(terra::vect(xy, crs = proj_string), "OGC:CRS84")),
    vapour     = function(xy) vapour::vapour_project(xy, "OGC:CRS84", proj_string))
}

## wrap: NaN where undefined, never error
safe <- function(f) function(m) {
  m <- matrix(as.numeric(m), ncol = 2)
  out <- tryCatch(f(m), error = function(e) matrix(NA_real_, nrow(m), 2))
  out <- matrix(as.numeric(out), ncol = 2)
  bad <- !is.finite(out[, 1]) | !is.finite(out[, 2]) | abs(out[, 1]) > 1e15 | abs(out[, 2]) > 1e15
  out[bad, ] <- NA_real_
  out
}
## inverse one point at a time so singular corners fail individually
pointwise <- function(f) function(m) {
  m <- matrix(as.numeric(m), ncol = 2)
  t(apply(m, 1, function(p) f(matrix(p, ncol = 2))))
}
forward <- safe(make_forward(SPILHAUS))
inverse <- pointwise(safe(make_inverse(SPILHAUS)))
```

## 2. The projection's domain

PROJ does not publish the square's half-width, so forward-project a dense
lon/lat grid and take the extent; then inverse-project the square's boundary
to trace the cut on the globe.

```{r domain}
square_half <- function(fwd, n = 721) {
  g <- as.matrix(expand.grid(lon = seq(-180, 180, length.out = n), lat = seq(-90, 90, length.out = (n + 1) %/% 2)))
  xy <- fwd(g)
  ceiling(max(abs(xy), na.rm = TRUE) / 1000) * 1000
}
HALF <- square_half(forward)

n <- 300; t <- seq(-1, 1, length.out = n); edge <- 0.996 * HALF
boundary <- rbind(cbind(t * edge, -edge), cbind(edge, t * edge), cbind(rev(t) * edge, edge), cbind(-edge, rev(t) * edge))
cut <- inverse(boundary)
corners <- rbind(`corner A (Asia)` = c(115, 30), `corner B (South America)` = c(-65, -30))

cat(sprintf("Half-width %s m; %d of %d boundary samples inverted (misses cluster at the singular corners)\n",
            format(HALF, big.mark = ","), sum(complete.cases(cut)), nrow(cut)))

op <- par(mar = c(3, 3, 2, 1))
plot(NA, xlim = c(-180, 180), ylim = c(-90, 90), asp = 1, xlab = "", ylab = "",
     main = sprintf("Where the Spilhaus cut falls on the globe (half-width %.3f Mm)", HALF / 1e6))
grid(); points(cut, pch = 16, cex = 0.4, col = "#b8342c")
points(corners, cex = 1.6, lwd = 1.5); text(corners, rownames(corners), pos = 4, cex = 0.8)
par(op)
```

## 3. Raster: bathymetry over `/vsicurl`

`gdalraster::warp()` is GDAL's warper with the same option vocabulary as
`gdalwarp`, so the target extent and size are stated explicitly rather than
estimated. The result is written to `/vsimem/` and read back as a matrix for
**ximage**; nothing touches disk and only the tiles GDAL needs come over the
wire.

```{r warp}
warp_square <- function(proj_string, half, npix, url = RASTER_URL, out = "/vsimem/spilhaus.tif") {
  t0 <- Sys.time()
  gdalraster::warp(url, out, t_srs = proj_string,
                   cl_arg = c("-te", -half, -half, half, half, "-ts", npix, npix,
                              "-r", "average", "-dstalpha", "-overwrite", "-q"))
  ds <- new(gdalraster::GDALRaster, out)
  on.exit(ds$close())
  v <- ds$read(band = 1, xoff = 0, yoff = 0, xsize = npix, ysize = npix, out_xsize = npix, out_ysize = npix)
  a <- ds$read(band = ds$getRasterCount(), xoff = 0, yoff = 0, xsize = npix, ysize = npix, out_xsize = npix, out_ysize = npix)
  m <- matrix(v, nrow = npix, ncol = npix, byrow = TRUE)   # row 1 = top of the image
  m[matrix(a, npix, npix, byrow = TRUE) == 0] <- NA
  list(z = m, res_m = 2 * half / npix, seconds = round(as.numeric(difftime(Sys.time(), t0, units = "secs")), 1),
       bands = ds$getRasterCount())
}
w <- warp_square(SPILHAUS, HALF, NPIX)
cat(sprintf("Warped %s to a %d x %d Spilhaus grid (%.1f km/px) in %s s\n",
            basename(RASTER_URL), NPIX, NPIX, w$res_m / 1000, w$seconds))
```

```{r render-raster}
bathy_cols <- grDevices::colorRampPalette(c("#08143f", "#123c7a", "#2a6fb5", "#6aa9d8", "#b9d9ea", "#e3f1f6"))(64)
render_raster <- function(z, half, add = FALSE) {
  ocean <- z; ocean[!is.na(z) & z >= 0] <- NA
  land  <- z; land[is.na(z) | z < 0] <- NA
  ximage::ximage(pmax(ocean, -7000), extent = c(-half, half, -half, half), asp = 1, col = bathy_cols,
                 zlim = c(-7000, 0), axes = FALSE, xlab = "", ylab = "", add = add)
  ximage::ximage(matrix(as.numeric(!is.na(land)), nrow(land)), extent = c(-half, half, -half, half),
                 zlim = c(0, 1), col = c(NA, "#d9d2c5"), add = TRUE)
}
op <- par(mar = c(0, 0, 2, 0), bg = "#d9d2c5")
render_raster(w$z, HALF)
title("Raster only (bathymetry), warped by GDAL via gdalraster", cex.main = 0.9)
par(op)
```

## 4. Vector: Natural Earth ocean polygons, as vertices

Read remotely with gdalraster's `GDALVector`, geometry returned as WKB, and
handed to **wk**. From there the polygons are just a coordinate table with
ring ids: transform the vertices, form segments between consecutive vertices
of the same ring, and drop the segments that are absurdly long in projected
space (those straddle the cut) or that are artefacts of the lon/lat
representation (both ends on the antimeridian or at a pole). Coastlines are
drawn as segments, not filled polygons.

```{r vector}
lyr <- new(gdalraster::GDALVector, OCEAN_URL)
lyr$returnGeomAs <- "WKB"
feat <- lyr$fetch(-1)
lyr$close()
geom_col <- which(vapply(feat, is.list, logical(1)))[1]  # the WKB column, whatever it is called
geom <- wk::wkb(feat[[geom_col]])
cat(sprintf("Ocean layer: %d feature(s), %s\n", length(geom), wk::wk_crs(geom)))

coast_segments <- function(geom, fwd, half, densify_deg = 0.5) {
  ## densify in lon/lat so the projected curves are smooth near the corners
  ## (wk has no densify; do it on the coordinate table)
  co <- wk::wk_coords(geom)
  key <- paste(co$feature_id, co$part_id, co$ring_id)
  dens <- do.call(rbind, lapply(split(co, key), function(r) {
    if (nrow(r) < 2) return(NULL)
    steps <- pmax(1L, ceiling(sqrt(diff(r$x)^2 + diff(r$y)^2) / densify_deg))
    x <- unlist(mapply(function(a, b, k) seq(a, b, length.out = k + 1)[-(k + 1)], r$x[-nrow(r)], r$x[-1], steps))
    y <- unlist(mapply(function(a, b, k) seq(a, b, length.out = k + 1)[-(k + 1)], r$y[-nrow(r)], r$y[-1], steps))
    data.frame(ring = r$ring_id[1] + 1e6 * r$feature_id[1], x = c(x, r$x[nrow(r)]), y = c(y, r$y[nrow(r)]))
  }))
  xy <- fwd(cbind(dens$x, dens$y))
  same_ring <- dens$ring[-1] == dens$ring[-nrow(dens)]
  step <- sqrt(diff(xy[, 1])^2 + diff(xy[, 2])^2)
  synthetic <- (abs(dens$x[-1]) > 179.99 & abs(dens$x[-nrow(dens)]) > 179.99) |
               (abs(dens$y[-1]) > 89.99 & abs(dens$y[-nrow(dens)]) > 89.99)
  ok <- same_ring & is.finite(step) & step < half / 8 & !synthetic
  i <- which(ok)
  cbind(x0 = xy[i, 1], y0 = xy[i, 2], x1 = xy[i + 1, 1], y1 = xy[i + 1, 2])
}
segs <- coast_segments(geom, forward, HALF)
cat(sprintf("%d coastline segments after projecting with %s and dropping cut-crossing segments\n", nrow(segs), ENGINE))
```

## 5. Points: landmarks, transformed directly

```{r points}
landmarks <- rbind(
  Hobart = c(147.33, -42.88), Tokyo = c(139.69, 35.69), Reykjavik = c(-21.94, 64.15),
  `Cape of Good Hope` = c(18.47, -34.36), `Bering Strait` = c(-169.0, 65.8),
  `Drake Passage` = c(-63.0, -59.0), `Point Nemo` = c(-123.39, -48.88),
  `Challenger Deep` = c(142.20, 11.37), `Puerto Rico Trench` = c(-66.5, 19.7),
  `Mid-Atlantic Ridge (Azores)` = c(-27.0, 38.0), `Macquarie Island` = c(158.94, -54.62),
  `Mawson Station` = c(62.87, -67.60), `Casey Station` = c(110.53, -66.28))
lm_xy <- forward(landmarks)
knitr::kable(data.frame(name = rownames(landmarks), lon = landmarks[, 1], lat = landmarks[, 2],
                        x = round(lm_xy[, 1]), y = round(lm_xy[, 2]), row.names = NULL),
             caption = "Landmarks in Spilhaus metres")
```

## 6. Composite

```{r composite, fig.width=11, fig.height=11}
composite <- function(z, half, segs, lm_xy, main) {
  op <- par(mar = c(0, 0, 3, 0), bg = "white")
  render_raster(z, half)
  segments(segs[, 1], segs[, 2], segs[, 3], segs[, 4], col = "#3a3a3a", lwd = 0.6)
  points(lm_xy, pch = 21, bg = "#f2b134", col = "black", cex = 1.1)
  text(lm_xy, rownames(landmarks), pos = 4, cex = 0.75, offset = 0.4)
  title(main, cex.main = 1)
  par(op)
}
proj_used <- provenance$proj[provenance$package == ENGINE]
composite(w$z, HALF, segs, lm_xy,
          sprintf("The world ocean in Spilhaus projection\nraster warped by GDAL (gdalraster); coastlines and points transformed by %s (PROJ %s)", ENGINE, proj_used))
```

## 7. Naming the thing: proj-string, WKT, and the missing authority code

Three renderings of "the same" projection. The bare proj-string exports to
WKT with the method named and *no parameters*: the numbers live in PROJ's
source. Spell the parameters out and the WKT carries them. **ESRI:54099** is
the only authority code you will find and it is a different, sqrt(2)-larger
square, which PROJ expresses as `+k_0=1.41421356237`. There is no EPSG code.

```{r wkt}
to_wkt <- function(x) tryCatch(gdalraster::srs_to_wkt(x, pretty = TRUE), error = function(e) paste("(", conditionMessage(e), ")"))
hobart <- function(x) tryCatch({ p <- safe(make_forward(x))(cbind(147.33, -42.88)); sprintf("%s, %s", format(round(p[1]), big.mark = ","), format(round(p[2]), big.mark = ",")) }, error = function(e) "transform failed")
renderings <- data.frame(
  rendering = c("bare proj-string", "explicit proj-string", "ESRI:54099", "alternative centre (section 8)"),
  input = c(SPILHAUS, SPILHAUS_EXPLICIT, "ESRI:54099", SPILHAUS_ALT))
renderings$hobart_xy <- vapply(renderings$input, hobart, "")
knitr::kable(renderings, caption = "Same projection, several names. hobart_xy is Hobart's projected coordinate under each.", row.names = FALSE)
```

```{r wkt-dump}
#| results: asis
fence <- strrep("`", 3)
for (i in seq_len(nrow(renderings))) {
  cat("\n#### WKT2 for", renderings$rendering[i], "\n\n", fence, "\n", to_wkt(renderings$input[i]), "\n", fence, "\n", sep = "")
}
```

## 8. Another member of the family

Same method, different parameters: a naive recentring on the Atlantic. The
square's edges now slice the Pacific. Spilhaus's contribution was the
parameter choice, not the projection.

```{r alt, fig.width=8, fig.height=8}
forward_alt <- safe(make_forward(SPILHAUS_ALT))
HALF_ALT <- square_half(forward_alt)
w_alt <- warp_square(SPILHAUS_ALT, HALF_ALT, max(400L, NPIX %/% 2L), out = "/vsimem/spilhaus_alt.tif")
segs_alt <- coast_segments(geom, forward_alt, HALF_ALT)
composite(w_alt$z, HALF_ALT, segs_alt, forward_alt(landmarks),
          sprintf("%s\n(half-width %.3f Mm vs %.3f Mm for the default)", SPILHAUS_ALT, HALF_ALT / 1e6, HALF / 1e6))
```

## Appendix: the same thing in one line each

Convenience layers hide which PROJ did the work. Each of these either
works or does not, purely according to the table in section 1.

```{r appendix-sf}
if (requireNamespace("sf", quietly = TRUE)) {
  x <- sf::st_transform(sf::st_sfc(sf::st_point(c(147.33, -42.88)), crs = "OGC:CRS84"), SPILHAUS)
  print(sf::st_coordinates(x)); cat("sf used PROJ", sf::sf_extSoftVersion()[["PROJ"]], "\n")
}
```

```{r appendix-terra}
if (requireNamespace("terra", quietly = TRUE)) {
  x <- terra::project(terra::vect(cbind(147.33, -42.88), crs = "OGC:CRS84"), SPILHAUS)
  print(terra::crds(x)); cat("terra used PROJ", terra::gdal(lib = "proj"), "\n")
}
```

## What to take from this

- On Linux the R packages usually agree with each other because they share
  one `libproj`; the question is whether the *system* PROJ is 9.6 or later.
  A stock Ubuntu LTS says no. The matrix in this repo's Actions shows what
  each runner image actually has.
- On macOS and Windows the CRAN binaries carry their own PROJ, and those
  versions move on CRAN's schedule.
- The primitives route (gdalraster for GDAL, wk for vertices, ximage for
  pixels) makes the dependency visible: every transform in this document is
  a call you can see, made by a package you can name.
- A proj-string is a recipe with hidden defaults. Cite the WKT with the
  parameters in it, and do not assume the one authority code you can find
  describes the map you meant.
